Reproducible Research: Peer Assessment 1
========================================

Loading and preprocessing the data
-------------------------------------

First, I have unzipped the data file.

```{r}
unzip("activity.zip")
```

Data contains three columns:

1. "steps" - integer number of steps taken during interval (contains NA values)

2. "date" - isoformat date of measurement

3. "interval" - interval number

Steps and interval will be automatically read as integer values, but date
needs to be processed.

I load it as character value, by disabling stringsAsFactor option
and then I convert it into a Date class.

```{r}
# read data
data <- read.csv("activity.csv")

# convert "date" to Date class
data$date <- as.Date(data$date)

# prepare a copy of data with only complete rows
completeData <- na.omit(data)

# preview data
str(data)
```

I also prepare copy of data with missing data omitted.

```{r}
# prepare a copy of data with only complete rows
completeData <- na.omit(data)

# preview data
str(completeData)
```


What is mean total number of steps taken per day?
-------------------------------------------------

I aggregate steps by date using only the complete measurements.

```{r}
# aggregate
stepsPerDay <- with(completeData, aggregate(steps, list(date), sum))
names(stepsPerDay) <- c("date", "steps")

# calculate mean and median
stepsMean <- mean(stepsPerDay$steps)
stepsMedian <- median(stepsPerDay$steps)

# preview data
str(stepsPerDay)
```

The *mean* number of steps per day is
`r format(stepsMean, scientific=F, nsmall=2)`
and *median* is `r format(stepsMedian, scientific=F, nsmall=2)`.

Let's see that on a histogram!

```{r}
hist(stepsPerDay$steps, col = "red", breaks = 20,
     xlab = "Total Number of Steps per Day",
     main = "Histogram of Total Number of Steps Taken Each Day")
abline(v=stepsMean, col="blue")
abline(v=stepsMedian, col="green")
rug(stepsPerDay$steps)
```



What is the average daily activity pattern?
-------------------------------------------

This time I aggregate mean steps by the interval number using only complete
measurements.

```{r}
dailyActivity <- with(completeData, aggregate(steps, list(interval), mean))
names(dailyActivity) <- c("interval", "steps")
```

Let's see that on time series plot!

```{r}
with(dailyActivity, plot(interval, steps, type='l',
                         main="Daily Activity Pattern"))
```

There seems to be a spike of activity in the morning.

```{r}
maxActivityInterval <- with(dailyActivity, interval[which.max(steps)])
```

The interval with maximum average number of steps taken is
`r maxActivityInterval`.



Imputing missing values
-----------------------

Note that there are a number of days/intervals where there are missing values
(coded as NA). The presence of missing days may introduce bias into some
calculations or summaries of the data.

```{r}
numberOfNA <- dim(data)[1] - dim(completeData)[1]
numberOfNAperDaySums <- (
    length(levels(factor(data$date))) # number of days present in date column
    - length(levels(factor(stepsPerDay$date))) # number of days with at least one measurement
)
```

Total number of rows with missing data is `r numberOfNA` and number of days
that do not have any measurement is `r numberOfNAperDaySums`.

I will try to fill the missing data using the average steps taken in given
interval. This seems to be a better strategy than using day's averages,
because some days have all the data missing and interval averages are more
informative.

```{r}
# merge per interval averages with incomplete data
filledData <- merge(data, dailyActivity, by="interval")

# construct steps column using original data where available
# and per interval averages, where data is missing
filledData$steps <- with(filledData, ifelse(is.na(steps.x), steps.y, steps.x))

# drop unneeded columns
filledData <- filledData[, names(data)]

# preview the data
str(filledData)
```

Now that we have completed data, let's repeat the calculation
of *mean* and *median* total number of steps taken per day.

I aggregate steps by date using only the complete measurements.

```{r}
# aggregate
stepsPerDay2 <- with(filledData, aggregate(steps, list(date), sum))
names(stepsPerDay2) <- c("date", "steps")

# calculate mean and median
stepsMean2 <- mean(stepsPerDay2$steps)
stepsMedian2 <- median(stepsPerDay2$steps)

# preview data
str(stepsPerDay2)
```

The *mean* number of steps per day is
`r format(stepsMean2, scientific=F, nsmall=2)`
and *median* is `r format(stepsMedian2, scientific=F, nsmall=2)`.

Let's see that on a histogram!

```{r}
hist(stepsPerDay2$steps, col = "red", breaks = 20,
     xlab = "Total Number of Steps per Day",
     main = "Histogram of Total Number of Steps Taken Each Day (adjusted data)")
abline(v=stepsMean2, col="blue")
abline(v=stepsMedian2, col="green")
rug(stepsPerDay2$steps)
```

It seems that *mean* and *median* have not changed much. Filling data using my
method resulted in adding `r numberOfNAperDaySums` days to the central bin.



Are there differences in activity patterns between weekdays and weekends?
-------------------------------------------------------------------------

In this analysis I will use the filled in data.

First, let's create a new factor variable in the dataset with two levels
– “weekday” and “weekend” indicating whether a given date is a weekday
or weekend day.

```{r}
# set locale to en_US to get English names of day
Sys.setlocale("LC_ALL", "en_US")

# create new column
filledData$weekPart <- factor(ifelse(
    weekdays(filledData$date) %in% c("Saturday", "Sunday"),
    "weekend", "weekday"))

# show that it is a factor
str(filledData)
```

Now, let's see a panel plot containing a time series plot of the 5-minute
interval (x-axis) and the average number of steps taken, averaged across
all weekday days or weekend days (y-axis).

```{r}
# prepare aggregate for the plot
weekPartActivity <- with(filledData, aggregate(steps, list(interval, weekPart), mean))
names(weekPartActivity) <- c("interval", "weekPart", "steps")

# preview data
str(weekPartActivity)
```

```{r}
library(lattice)
xyplot(steps ~ interval | weekPart, data=weekPartActivity,
       type="l", layout=c(1, 2))
```

It seems that weekend and weekday activity patterns are different.
On weekdays there is a larger spike in the morning, but lower overall activity
in the rest of the day. Weekend activity is more evenly spread out throughout
the day.
