Reproducible Research: Peer Assessment 1
========================================

Loading and preprocessing the data
-------------------------------------

First, I have unziped the data file.

```{r}
unzip("activity.zip")
```

Data contains three columns:

1. "steps" - integer number of steps taken during interval (contains NA values)

2. "date" - isoformat date of measurement

3. "interval" - interval number

Steps and interval will be automatically read as integer values, but date
needs to be processed.

I load it as character value, by disabling stringsAsFactor option
and then I convert it into

```{r}
# read data
data <- read.csv("activity.csv")
# convert "date" to Date class
data$date <- as.Date(data$date)
# prepare a copy of data with only complete rows
completeData <- na.omit(data)
```


What is mean total number of steps taken per day?
-------------------------------------------------

I aggregate steps by date using only the complete measurements.

```{r}
stepsPerDay <- with(completeData, aggregate(steps, list(date), sum))
names(stepsPerDay)[2] <- "steps"
stepsMean <- mean(stepsPerDay$steps)
stepsMedian <- median(stepsPerDay$steps)
```

The mean number of steps per day is
`r format(stepsMean, scientific=F, nsmall=2)`
and median is `r stepsMedian`.

Let's see that on a histogram!

```{r}
hist(stepsPerDay$steps, col = "red", breaks = 20,
     xlab = "Total number of steps per day",
     main = "Histogram of total number of steps taken each day")
abline(v=stepsMean, col="blue")
abline(v=stepsMedian, col="green")
```



What is the average daily activity pattern?
-------------------------------------------



Imputing missing values
-----------------------



Are there differences in activity patterns between weekdays and weekends?
-------------------------------------------------------------------------
