---
title: "Weight Lifting Exercise Prediction"
output: html_document
---

Load and Clenup the Data
------------------------

First, I load training and testing datasets.

```{r}
training <- read.csv('pml-training.csv')
testing <- read.csv('pml-testing.csv')
```

I select the columns that are usable for constructing the prediction model.
I chose _roll_, _pitch_, _gyros{x,y,z}_, _accel{x,y,z}_ and _magnet{x,y,z}_
for each body part, that is _belt_, _arm_, _dumbbell_ and _forearm_.

```{r}
features <- c(
    # belt
    'roll_belt', 'pitch_belt', 'yaw_belt',
    'gyros_belt_x', 'gyros_belt_y', 'gyros_belt_z',
    'accel_belt_x', 'accel_belt_y', 'accel_belt_z',
    'magnet_belt_x', 'magnet_belt_y', 'magnet_belt_z',
    # arm
    'roll_arm', 'pitch_arm', 'yaw_arm',
    'gyros_arm_x', 'gyros_arm_y', 'gyros_arm_z',
    'accel_arm_x', 'accel_arm_y', 'accel_arm_z',
    'magnet_arm_x', 'magnet_arm_y', 'magnet_arm_z',
    # dumbbell
    'roll_dumbbell', 'pitch_dumbbell', 'yaw_dumbbell',
    'gyros_dumbbell_x', 'gyros_dumbbell_y', 'gyros_dumbbell_z',
    'accel_dumbbell_x', 'accel_dumbbell_y', 'accel_dumbbell_z',
    'magnet_dumbbell_x', 'magnet_dumbbell_y', 'magnet_dumbbell_z',
    # forearm
    'roll_forearm', 'pitch_forearm', 'yaw_forearm',
    'gyros_forearm_x', 'gyros_forearm_y', 'gyros_forearm_z',
    'accel_forearm_x', 'accel_forearm_y', 'accel_forearm_z',
    'magnet_forearm_x', 'magnet_forearm_y', 'magnet_forearm_z'
    )
training <- training[, c('classe', features)]
testing <- testing[, features]
```

Build Prediction Model
----------------------

I have chosen the Support Vector Machine with Radial Basis Function Kernel (SVM)
as a prediction engine.
I have used 10 fold cross-validation to estimate out-of-sample error.

```{r}

library(caret)
set.seed(123456)
fitControl <- trainControl(
    ## 10-fold CV
    method = "cv",
    number = 10,
)
fit <- train(
    classe ~ .,
    data = training,
    method = "svmRadial",
    preProc = c("center", "scale"),
    tuneLength = 8,
    trainControl = fitControl
)
fit
```

Final model out-of-sample error estimation is `r 1-getTrainPerf(fit)$TrainAccuracy` as calculated in:
```{r}
getTrainPerf(fit)
```

```{r}
out.of.sample.error <- 1 - getTrainPerf(fit)$TrainAccuracy
out.of.sample.error
```

Predicting the Test Data
------------------------

I classify the test data using my model:

```{r}
prediction <- as.character(predict(fit, testing))
prediction
```

I save prediction to txt files:

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(prediction)
```