---
title: "Motor Trends: Automatic or a Stick? It doesn't matter."
output: pdf_document
---

Summary
-------

We take a look at some historical data (the 1974 Motor Trend US magazine car
road tests) and try to answer what car characterstics influence fuel confumption
(measured in Miles per Gallon). We will take a closer look on the relation
between transmission type (auto or stick) and car's MPG and observe that if
we include other important variables like car's weight and number of cylinders
in car's engine, then the relation between transmission type and fuel consumption
is insignificant.

Exploratory Analysis
--------------------

First, we take a look at the data to see if there are some obvious correlations.

```{r}
data(mtcars)
pairs(mtcars)
```

Let's see correlation table for completeness:
```{r}
cor(mtcars)
```

MPG seems to be highly correlated with number of cylinders (cyl), displacement
(disp), horsepower (hp) and weight (wt), so we will use those variables in our
regression models. Transmission type doesn't seem to be highly correlated, but
we will include it in our model to be able to answer our research question.


Models
------

We will create a series of regression models of increasing complexity.
We will start with the simplest one, where the MPG is being explained only
by the transmission type.

```{r}
fit <- lm(mpg ~ am, data = mtcars)
summary(fit)
```

This simple model is highly interpretable - intercept coeficient is
`r coef(fit)[1]`, which means that a car with automatic transmission
(am == 0) is expected to drive `r coef(fit)[1]` miles per gallon and am
coefficient is `r coef(fit)[2]`, which means that a car with manual
transmission (am == 1) is expected to drive `r coef(fit)[2]` more miles per gallon.
All p-values are much lower than 0.05, so the model seems to be strong, but
the variation explained (R-squared) is only `r summary(fit)$r.squared`, so we
need to include more variables to have a more complete view on cars' MPG.

Let's see the relation beetween transmission type and MPG on a plot.
```{r}
plot(mtcars$am, mtcars$mpg, xlab="Transmission Type", ylab="MPG", main="MPG vs Transmission Type")
abline(fit)
```


We will try to create a series of more complex models to get a better view on
fuel consumption.
We will regress on variables centered on mean or minimal value of those
variables, so the intercept would remain interpretable - it doesn't have sense
to have an intercept that is a MPG value for car with zero horsepower,
zero weight and zero cylinders, but it does have sense to have an intercept as
a value for car with mean horsepower (`r mean(mtcars$hp)` HP) and weight
(`r mean(mtcars$wt)*1000` lb) and minimal number of cylinders
(`r min(mtcars$cyl)` cylinders).

```{r}
mtcars$cyl.2 <- mtcars$cyl - min(mtcars$cyl)
mtcars$disp.2 <- mtcars$disp - mean(mtcars$disp)
mtcars$hp.2 <- mtcars$hp - mean(mtcars$hp)
mtcars$wt.2 <- (mtcars$wt - mean(mtcars$wt)) * 1000 # also scale to lb
```

We add variables to regressin model in order of correlation and test their
significance using ANOVA test:
```{r}
fit2 <- lm(mpg ~ am + wt.2, data = mtcars)
fit3 <- lm(mpg ~ am + wt.2 + cyl.2, data = mtcars)
fit4 <- lm(mpg ~ am + wt.2 + cyl.2 + disp.2, data = mtcars)
fit5 <- lm(mpg ~ am + wt.2 + cyl.2 + disp.2 + hp.2, data = mtcars)
anova(fit, fit2, fit3, fit4, fit5)
```

As we can see, adding weight and cylinder information results with significanlty
better models (p-value < 0.05) and adding displacement and horse power does not.
In result, we use fit3 model using transmission type, weight and number
of cylinders as our final model.

Let's see it's details:
```{r}
summary(fit3)
```

The intercept intercept coeficient is
`r coef(fit)[1]`, which means that a car with automatic transmission
(am == 0), `r min(mtcars$cyl)` cylinder engine and average weight is
expected to have fuel efficiency of `r coef(fit3)[1]` miles per gallon.

The transmission type coefficient is only `r coef(fit3)[2]` and in the first
model it was `r coef(fit)[1]`, which means that the newly added variables
(weight and number of cylinders were confounding the transmission type variable).
The p-value of this coefficient is very high (0.89334), so it means that after
taking an account of car's weight and number of cylinders the relation between
MPG and transmission type diminishes.

The final model seems to be strong with R-squared value of
`r summary(fit3)$r.squared' and resudual plot without anomalies:

```{r}
plot(fit3, which = 1)
```

Conclusion
----------

Is an automatic or manual transmission better for MPG?
On the surface, there is a difference, but if you look in depth
and take an account for car's weight and engine's number of cylinders,
the transmission type of your car doesn't matter.
